---
title: "Pattern of association between registered covid cases and regsitered number of death"
subtitle: "Assignment for Data analysis 2 and Coding with R"
author: "Viktória Mészáros"
date: "10-26-2020"
output: 
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.asp = 0.8, fig.width = 7, out.width = "90%" )
```
# Executive summary
In this assignment I looked for the association between confirmed Covid cases and the number of deaths due to Covid.
After checking several different possible models I chose a log-log weighted linear regression model, using population as weights as the best representation of the pattern of association between the two variables. With this model we can make the interpretation that if the number of cases is 10% higher, the number of deaths is 9.5% higher, on average.




## Aim of analysis

The aim of this analysis was to look at the pattern of association between registered Covid cases and registered number of death due to Covid from data collected from the start of the pandemic, until October 26, 2020. For this I used two separate data sources. The first data set was a [Covid-19] (https://github.com/CSSEGISandData/COVID-19) data  which was collected by Center for Systems Science and Engineering (CSSE) at Johns Hopkins University. The other was population data for 2019 from the World Bank's site. 

Variables in my analysis:

- Outcome variable: Number of registered death
- Explanatory variable: Number of registered cases
- Weight: Population (I will use the population for each county to calculate a weighed model (reg 8))

*All variables are measured as number of people.*
  
## Data cleaning 
Before beiigng able to do analysis on these data, I had to do some cleaning.

*Covid data:*

- Dropped several variables which were not relevant for my analysis (only kept State, Country, Confirmed, Death)
- To have a country level data, so I summarized the regions for each country

*Population data:*

- Removed all non-country observations

I merged the two cleaned data tables and also filtered out the variables where population, cases or the number od death was missing. 

**My research question is:**
  "How does the change in the amount of Covid cases effect the number of deaths due to Covid?"

```{r, include=F, message=F, error=F, warning= F}
# Clear memory
rm(list=ls())

# Packages to use
library(tidyverse)
require(scales)
library(lspline)
library(estimatr)
library(texreg)
library(ggthemes)
library(xtable)
library(moments)
library(kableExtra)
library(car)

# Call the data from github
my_path <- "~/Documents/CEU/Fall_semester/Data_analysis_2/Assignment/Data/"
df <- read_csv(paste0(my_path, 'Clean/covid_pop_10_26_2020_clean.csv'))

```

## Quick check on all of my varibles
From the graph we can see that all the distributions for our variables are skewed to the right. Before moving on, we should have a look at the extreme values.

```{r, message= F, warning= F}
df %>%
  keep(is.numeric) %>% 
  gather() %>% 
  ggplot(aes(value)) +
  facet_wrap(~key, scales = "free") +
  geom_histogram(fill = "deepskyblue4")+
  theme_clean() + 
  scale_fill_gdocs()
```


## Extreme values
When I take a closer look on the potential extreme values, I realize the really large numbers for cases and deaths belong to the United States, India and Brazil meaning these are not due to error, but important observations for my analysis, so I do not exclude them. The issue is similar for population. These the large values are China and India which are not extreme values either. There are no unknown values in any variables, as I excluded them during the cleaning phase. There are 14 countries in the data, where the number of deaths were zero. Together these represent 0.004% of the cases and 0.3% of the total population. As the aim of my analysis is to find a pattern between the number of cases and deaths, I decided to exclude those countries from the data. 
```{r, include = F}
# CASES
df %>% 
  select(c( country, cases)) %>% 
  arrange(-cases) %>% 
  head(5)
# The extreme values for cases are not errors, but the number of cases in the United States (8.7M),
# India (7.9M) and Brazil(5.4M). I won't exclude these, as they are important observations in my analysis.  

df %>% 
  filter(cases < 1000)
# There are 31 observations with less than 1000 Covid cases registered.
# There are no observations with NA or zero values, as we excluded them during the data cleaning phase

# DEATH
df %>% 
  select(c( country, death, cases)) %>% 
  arrange(-death) %>% 
  head(10)
# The four extreme values visible on the graph are United States(2.3M), Brazil(1.6M), India(1.2M) and Mexico(0.9M).
# The top three are the same countries for cases and deaths as well, which totally makes sense, as where the number of 
# infections are higher, the number of deaths are higher as well. We don't exclude values as the are not extreme.

df %>% filter(death == 0)
# There are 12 countries in where no one died due to Covid. In all of these countries the number of cases
# were below 500. All in all 1890 cases happened in these countries, compared to total of 43M cases (0.004%).
# As the aim of my analysis is to find if a pattern exist between the number of cases and deaths, 
# I will exclude these countries as they has no meaning for my analysis, and they represent a really small portion of 
# observed cases.

df <- df %>% filter(death > 0)

# POPULATION 
df %>% 
  select(c(country, population)) %>% 
  arrange(-population ) %>% 
  head(10)
# If we check population data, we see that the large values are India and China so we don't exclude any values here either. There are some countries with unknown populations, but as I only use population as a weight and don't use it in several models, I do not exclude these observations.

summary( df )
```
## Scaling
For easier interpretation and also because there are 52 observations (our of 175) where the number of deaths were below 100, I decided to leave the values as people and do not transform them to hundred/thousand or million people. To remain consistent, I did not do any scaling for population and cases either, so all my variables are measured the same way, showing number of people.

## Distributions for the x and y variables
Both from the histogram and the statistics
```{r, echo= F, message=F, warning= F}
df %>%
  select(c(cases, death)) %>% 
  gather() %>% 
  ggplot(aes(value)) +
  facet_wrap(~key, scales = "free") +
  geom_histogram(fill = "deepskyblue4")+
  theme_clean() + 
  scale_fill_gdocs()

cases_sum <- df %>% 
  summarise(mean = mean(cases),
            median   = median(cases),
            min      = min(cases),
            max      = max(cases),
            std      = sd(cases),
            skew     = round(skewness(cases), digit=4),
            numObs   = sum( !is.na( cases ) ) )
cases_tit <- data.frame(variable = "Registered Covid cases")
cases_sum <- cbind(cases_tit, cases_sum)

death_sum <- df %>% 
  summarise(mean = mean(death),
            median   = median(death),
            min      = min(death),
            max      = max(death),
            std      = sd(death),
            skew     = round(skewness(death), digit=4),
            numObs   = sum( !is.na( death ) ) )
death_tit <- data.frame(variable = "Registered deaths due to Covid")
death_sum <- cbind(death_tit, death_sum)

var_summary <- cases_sum %>% add_row( death_sum )

var_summary %>% 
  kable("html") %>%
  kable_material(font_size = 12)
```

## Possible ln() transformations
Basic scatter-plot showing the number of cases and the number of deaths - **level-level** model without scaling
```{r, echo= F, message= F, warning= F}
ggplot( df , aes(x = cases, y = death)) +
  geom_point() +
  geom_smooth(method="loess", color = "deepskyblue4")+
  theme_clean() +
  labs(x = "Number of registered cases",y = "Number of deaths") +
  annotate("text", x = 8700000, y = 219000, label = "United States", size=4)+
  annotate("text", x = 5400000, y = 152000, label = "Brazil", size=4)+
  annotate("text", x = 7920000,  y = 114000, label = "India", size=4)+
  annotate("text", x = 890000,  y = 84000, label = "Mexico", size=4)
```

Potential **level-log** model scaling for the number of cases 
```{r, echo= F, message= F, warning= F}

ggplot( df , aes(x = cases, y = death)) +
  geom_point() +
  geom_smooth(method="loess", color = "deepskyblue4")+
  theme_clean()+
  labs(title= '', x = "Number of registered cases (ln scale)",y = "Number of deaths") +
  scale_x_continuous( trans = log_trans(),  breaks = c(1,10,100,1000,10000, 100000, 500000) ) +
  annotate("text", x = 8700000, y = 219000, label = "United States", size=4)+
  annotate("text", x = 5400000, y = 152000, label = "Brazil", size=4)+
  annotate("text", x = 7920000,  y = 114000, label = "India", size=4)+
  annotate("text", x = 890000,  y = 84000, label = "Mexico", size=4)
```

Potential **log-level** model scaling for the number of deaths 
```{r, echo= F, message= F, warning= F}
ggplot( df , aes(x = cases, y = death)) +
  geom_point() +
  geom_smooth(method="loess", color = "deepskyblue4")+
  theme_clean()+
  labs(x = "Number of registered cases",y = "Number of deaths (ln scale)") +
  scale_y_continuous( trans = log_trans(),  breaks = c(1,10,100, 500, 5000,50000) ) +
  annotate("text", x = 8700000, y = 450000, label = "United States", size=4)+
  annotate("text", x = 5400000, y = 90000, label = "Brazil", size=4)+
  annotate("text", x = 7920000,  y = 70000, label = "India", size=4)
```

Potential **log-log** model scaling both the number of cases AND deaths
```{r, echo= F, message= F, warning= F}
ggplot( df , aes(x = cases, y = death))  +
  geom_point() +
  geom_smooth(method="loess", color = "deepskyblue4")+
  theme_clean()+
  labs(x = "Number of registered cases (ln scale)",y = "Number of deaths (ln scale)") +
  scale_x_continuous( trans = log_trans(),  breaks = c(1,10,100, 500, 5000, 50000) )+
  scale_y_continuous( trans = log_trans(), breaks = c(1,10,100, 500, 5000, 50000) )

```

Based on the graphs above I decided to take the log of the number of cases and the number of deaths as well. From a substantive view the number of cases and even deaths can increase by hundreds or thousand each day and so taking them as percentage can make interpretation easier (the variables are affected in a multiplicative way), also as the population of different countries differ significantly, a relative difference could be used better than a unit change. As a statistical reason, the distribution of the variables are skewed to the right, so with a ln() transformation it becomes more symmetric and closer to normal. (To take logs it is important that we only have positive values and as we dropped all zero values it is true for our data.)
```{r, include= F}
############################
#       Conclusions       ##
############################

#   1) taking log of the number of cases is needed, but still non-linear pattern in data (it is important that we excluded observations with 0 cases) 
#      as know we only have positive values, so we can take the logs)
#       - Substantive: Level changes is harder to interpret and it gives better understanding to get 
#                      if the number of cases increase by 1% what happens with the number of deaths, 
#                      rather than to see in case 1 more people gets the virus.
#       - Statistical: Due to the different population sizes of the countries, the number of cases has a long right tail (is skewed to the right), 
#                      with the ln() transformation its distribution will be closer to normal 
#   2) using only the number of deaths is possible, but need to model the non-linearity in the data
#       
#   3) taking log of the number of deaths is making the association close to linear!
#       - Substantive: it is likely affected in multiplicative ways, percentage changes are easy to interpret as the number of deaths can have highly #                        different values across countries, so relative percentage differences make comparison easier
#       - Statistical: the number of deaths also has a distribution skewed to the right, so it is feasible to make a ln() transformation
#
  
# Take Log of number of confirmed cases and log number of deaths
df <- df %>% mutate( ln_cases = log( cases ),
                     ln_death = log( death ) )

# Add powers of the variable(s) to the dataframe:
df <- df %>% mutate( ln_cases_2 = ln_cases^2,
                     ln_cases_3 = ln_cases^3)
```


## Pattern of association
I choose the weighted linear regression using population as weights as my preferred model. This was due to that I think in this case it makes more sense to analyze relative differences and the R^2 for this model with 92.7% is higher compared to any other (89% is the highest among other models). This model captures patterns the best and it can be interpret in a meaningful way. 
The interpretation of the intercept (-3) is not really meaningful in this case. Theoretically it shows that if ln(Number of cases) is zero, so when the number of cases is 1, the average ln(number of deaths) is -3. The slope parameter is more meaningful, showing that if the number of cases is 10% higher, the number of deaths is 9.5% higher, on average.
```{r , message=FALSE, echo= F}

reg8 <- lm_robust(ln_death ~ ln_cases, data = df , weights = population)

ggplot(data = df, aes(x = ln_cases, y = ln_death)) +
  geom_point(data = df, aes(size=population),  color = 'deepskyblue4', shape = 16, alpha = 0.6,  show.legend=F) +
  geom_smooth(aes(weight = population), method = "lm", color='deeppink4')+
  scale_size(range = c(1, 15)) +
  theme_clean()+
  labs(title = 'Regression 8', subtitle = 'Weighted-OLS model using ln(number of deaths)',
       x = "ln(Number of registered cases)", y = "ln(Number of deaths)" )+
  annotate("text", x = 11.4, y = 8.5, label = "China", size=4)+
  annotate("text", x = 15.5, y = 12.7, label = "United States", size=4)+
  annotate("text", x = 15.9,  y = 11.7, label = "India", size=4)

summary( reg8 )
```


## Test for the slope parameter
I use hypothesis testing on the association between confirmed cases and number of deaths. The null hypothesis is that beta equals zero, while my alternative hypothesis is that it is not zero. Let's use a 99% confidence interval. The t-value of the test is 15.65 which is out of the 2.6 (99% CI). We can also look at the p-value which is close to zero meaning that the possibility of a false positive error is almost 0%. As a result we can reject the H0 hypothesis that the beta coefficient on log confirmed cases is equal to zero. The correlation between the two variables seems to significant. 

```{r, echo=FALSE, message=FALSE, results='asis'}


linearHypothesis( reg8 , "ln_cases = 0")

```

## Analyze residuals
With looking at the residuals we can see which counties had higher or lower ln(number of death) compared what is expected based on out model. The best five countries, where the ln(deaths) is much lower compared to what was expected were Singapore, Burundi, Sri Lanka, Qatar and Liechtenstein. this can mean that these countires manage this pandemic better, compared to other, they have better healthcare system, or made more precautions. On the other hand, the countries, who did the worst were Yemen, Mexico, Ecuador, Italy and Iran. The reason behind can be the bad healthcare system in these countries or the lack of social distancing and other protective measure.

```{r , echo= F }

# Get the predicted y values from the model
df$reg8_y_pred <- reg8$fitted.values
df$reg8_y_pred_num <- round(exp(df$reg8_y_pred),2)

# Calculate the errors of the model
df$reg8_res <- df$ln_death - df$reg8_y_pred 
df$reg8_res_num <- df$death-df$reg8_y_pred_num

# Find countries with largest negative errors
df %>% top_n( -5 , reg8_res ) %>% 
      select( country , ln_death, reg8_y_pred , reg8_res ) %>% 
  arrange(reg8_res)

# Find countries with largest positive errors
df %>% top_n( 5 , reg8_res ) %>% 
  select( country , ln_death, reg8_y_pred , reg8_res ) %>% 
  arrange(-reg8_res)


```

``



## Conclusion

In this project I investigated the pattern of association between the number of registered Covid cases and the number of deaths due to this disease. I have found a strong correlation between these variables, seeing that if the number of cases are 10% higher, the number of deaths are also almost 10% (9.5% exactly) higher, on average

My analysis could have been strengthened by incorporating some other variables, such as the average age of people died or even a time frame somehow. As the data shows the cases as total until 26th of October, it would be interesting to see how they change with time in the different countries due to governmental rules or other issues. Another weakening factor can be in my analysis that not all the countries report the number of deaths the same time, and there are several cases where the reason behind a death is not purely due to covid. Taking more detaileds into consideration would make the analysis even stronger.

## Appendix I.- All regression models
```{r, echo= F}
# Do the regressions
#
#
# Models with actual number of deaths
#
# First model
reg1 <- lm_robust( death ~ ln_cases , data = df , se_type = "HC2" )
summary( reg1 )

ggplot( data = df, aes( x = ln_cases, y = death ) ) + 
  geom_point( color="deepskyblue4", alpha = 0.8) +
  geom_smooth( method = lm , color = 'deeppink4' )+
  theme_clean()+
  labs(title = 'Regression 1', subtitle = 'Level-log  model using number of deaths',
       x = "ln(Number of registered cases)", y = "Number of deaths" )

# Second model
reg2 <- lm_robust( death ~ ln_cases + ln_cases_2 , data = df )
summary( reg2 )

ggplot( data = df, aes( x = ln_cases, y = death ) ) + 
  geom_point( color= 'deepskyblue4', alpha = 0.8) +
  geom_smooth( formula = y ~ poly(x,2) , method = lm , color = 'deeppink4' )+
  theme_clean()+
  labs(title = 'Regression 2', subtitle = 'Level-log quadratic model using number of deaths',
       x = "ln(Number of registered cases)", y = "Number of deaths" )

# Third model
reg3 <- lm_robust( death ~ ln_cases + ln_cases_2 + ln_cases_3, data = df )
summary( reg3 )

ggplot( data = df, aes( x = ln_cases, y = death ) ) + 
  geom_point( color='deepskyblue4', alpha = 0.8) +
  geom_smooth( formula = y ~ poly(x,3) , method = lm , color = 'deeppink4' )+
  theme_clean()+
  labs(title = 'Regression 3', subtitle = 'Level-log cubic model using number of deaths',
       x = "ln(Number of registered cases)", y = "Number of deaths" )

# Fourth model 
# Regression with piecewise linear spline:

cutoff_1 <- 480000

ln_cutoff_1 <- log(cutoff_1)

# Use simple regression with the lspline function

reg4 <- lm_robust(death ~ lspline( ln_cases , ln_cutoff_1 ), data = df )
summary( reg4 )

ggplot( data = df, aes( x = ln_cases, y = death ) ) + 
  geom_point( color='deepskyblue4', alpha = 0.8) +
  geom_smooth( formula = y ~ lspline(x, ln_cutoff_1) , method = lm , color = 'deeppink4' )+
  theme_clean()+
  labs(title = 'Regression 4', subtitle = 'Piecewise linear spline',
       x = "ln(Number of registered cases)", y = "Number of deaths" )


# Models with ln(death)
#
# Fifth model
reg5 <- lm_robust( ln_death ~ ln_cases , data = df )
summary( reg5 )

ggplot( data = df, aes( x = ln_cases, y = ln_death ) ) + 
  geom_point( color='deepskyblue4', alpha = 0.8) +
  geom_smooth( method = lm , color = 'deeppink4' ) +
  theme_clean()+
  labs(title = 'Regression 5', subtitle = 'Log-log model using ln(number of deaths)',
       x = "ln(Number of registered cases)", y = "ln(Number of deaths)" )

# Sixth model
reg6 <- lm_robust( ln_death ~ ln_cases + ln_cases_2 , data = df )
summary( reg6 )

ggplot( data = df, aes( x = ln_cases, y = ln_death ) ) + 
  geom_point( color='deepskyblue4', alpha = 0.8) +
  geom_smooth( formula = y ~ poly(x,2), method = lm , color = 'deeppink4' ) +
  theme_clean()+
  labs(title = 'Regression 6', subtitle = 'Log-log quadratic model using ln(number of deaths)',
       x = "ln(Number of registered cases)", y = "ln(Number of deaths)" )

# Seventh model
reg7 <- lm_robust( ln_death ~ ln_cases + ln_cases_2 + ln_cases_3, data = df )
summary( reg7 )

ggplot( data = df, aes( x = ln_cases, y = ln_death ) ) + 
  geom_point( color='deepskyblue4', alpha = 0.8) +
  geom_smooth( formula = y ~ poly(x,3), method = lm , color = 'deeppink4' ) +
  theme_clean()+
  labs(title = 'Regression 7', subtitle = 'Log-log cubic model using ln(number of deaths)',
       x = "ln(Number of registered cases)", y = "ln(Number of deaths)" )


# Eight model
# Weighted-OLS: use reg6 setup and weight with population

reg8 <- lm_robust(ln_death ~ ln_cases, data = df , weights = population)
summary( reg8 )

ggplot(data = df, aes(x = ln_cases, y = ln_death)) +
  geom_point(data = df, aes(size=population),  color = 'deepskyblue4', shape = 16, alpha = 0.6,  show.legend=F) +
  geom_smooth(aes(weight = population), method = "lm", color='deeppink4')+
  scale_size(range = c(1, 15)) +
  theme_clean()+
  labs(title = 'Regression 8', subtitle = 'Weighted-OLS model using ln(number of deaths)',
       x = "ln(Number of registered cases)", y = "ln(Number of deaths)" )+
  annotate("text", x = 11.4, y = 8.5, label = "China", size=4)+
  annotate("text", x = 15.5, y = 12.7, label = "United States", size=4)+
  annotate("text", x = 15.9,  y = 11.7, label = "India", size=4)
  
```

## Appendix II. - Summary Table
```{r, echo= F}
#######################################
##    Creating model summary table   ##
#######################################
# 

data_out <- "~/Documents/CEU/Fall_semester/Data_analysis_2/Assignment/Out/"
htmlreg( list(reg1 , reg2 , reg3 , reg4 , reg5 , reg6 , reg7, reg8),
         type = 'html',
         custom.model.names = c("Number of deaths - linear","Number of deaths - quadratic","Number of deaths - cubic", "Number of deaths - PLS",
                                "ln(death) - linear","ln(death) - quadratic","ln(death) - cubic", "ln(death) - WOLS"),
         caption = "Modelling the number of deaths and the confirmed cases for Corona virus by countries",
         file = paste0( data_out ,'model_comparison.html'), include.ci = FALSE)

```

          



